Supplementary Information

### 2020-06-06 doc/manuscript/MEE_manuscript.Rmd
update manuscript
mainly wrt size
also minor fig updates/ cleanup

**Characterizing the influence of slowly-changing environtmental drivers is very challenging and uncertain.**
We want to know how changes in CO2 and atmospheric deposition are influencing tree growth.
A signficant challenge arises when tree-rings are used to characterize the influence of slowly changing environmental variables--climate, CO2, atmospheric deposition--on tree growth rates (REFS).
In these cases, a variety of methods are used to account for simultaneous changes in tree size, including... (DENDRO_REFS)
None of these methods is fully satisfactory; rather, it is necessary to *simultaneously* account for the influences of changing tree size and environmental drivers.

**Here, develop a new method that allows simultaneous consideration of the effects of tree size, objectively determined principle climate drivers, and other environemntal drivers on annual tree growth.** 
This allows us to ask:
(1) What are the most important climate drivers of annual growth, and over which time windows?
(2) What is the shape of the relationship between annual growth and climate drivers?
(3) How do radial growth and aboveground biomass increments ($\Delta AGB$) vary with DBH?
(4) Are there interactions between DBH and climate drivers common?
(5) How is growth simultaneously influenced by DBH, climate, and other environemental drivers (CO2, atmospheric deposition)?

@@

First, we identified the most important climate driver for each of three categories of variables: a temperature group (mean, min, and max temperature; PET), a precipitation group (precipitation, number of days with precipitation), and a group linked to temperature variability and cloud cover (daily temperature range, percent cloud cover, and again PET). 
These groups were defined based on the dependencies of these variables on one another in the CRU database (Harris et al. 2014, [2020](https://rdcu.be/b3U3e)), as well as climatological and biological considerations.
Our first step was detrending to remove the influence of all non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), which is essential for identifying climatic drivers (DENDRO_REFS).
Specifically, we used a GAM model to fit a spline to individual tree growth records ($log[\Delta r]$ or $log[\Delta AGB]$), thereby producing residuals.
We note that an an attempt at the analysis without detrending failed to produce results consistent with biological expectations and results obtained via traditional dendrochronological methods, and generally failed to identify significant climate drivers.

We then used the *climwin* package in R (REF) [(van de Pol et. al. 2016)](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12590) to identify the most significant climate driver and time window for each of the three climate variable groups. 
*Climwin* searches through multiple climate variables and over a wide range of time frames to identify which are most strongly correlated to the biological variable of interest--in this case, residuals of $log[\Delta r]$ or $log[\Delta AGB]$.
Within *climwin*, we specified a mixed effects model using species and tree identity as random effect: `residual ~ [climate] + (1 | sp) + (1 | treeID)`. 
Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of time periods, at monthly resolution, starting in the previous May and ending in August of the current year.
We note that analyzing all species together yields the most significant climate drivers across the full set of cores from each site (our goal here), whereas identification of the top climate drivers for individual species would be optimized by analyzing each species separately.
*Climwin* can check for linear and quadratic relationships, and does k-fold cross-validation in its computation of AIC to guard against over-fitting (CLIMWIN_REF). 
We specified quadratic relationships only because (i) quadratic relationships are more consistent with known biological mechanisms (see Introduction), (ii) preliminary tests revealed that quadratic fits usually had lower AIC, and when not there tended to be little difference in AIC and the curve would approximate a straight line, (iii) modeling only quadratic improves coding efficiency and speeds up the analysis process.
For each group of candidate climate variables, we move forward with the best variable over the time window identified by climwin as a candidate climate variable for the multivariate models. 
When PET came out as the best variable in both temperature and cloud cover groups, there were only two candidate variables for the GAM. 
We checked for colinearity among the selected climate variables and removed any variable with a variance inflation factor > 10.

@@

A challenge to this system arose for the site undergoing most rapid climate change: Scotty Creek. 
There, [temperatures have increased by X $^\circ$ over X years]..., resulting in negative growth trends in basal area index (BAI) starting around 1950 and significant growth declines since 1970 in `r round(249/(249+69+127)*100, digits=0)`% of trees (Sniderhan & Baltzer 2016).
Problematically, correlating tree growth residuals from which climate-driven trends had been removed against the climate signal with a strong directional trend would not necessarily identify the most signficant climate drivers.
(**[ISSUE #25 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25): How do we solve this? Any possibility besides subjectively selecting the drivers? Maybe try analysis without detrending**)

@@

Second, we combined DBH, climate, and atmospheric pollution data in a GLS.

@@

**Our process picked out similar climate drivers to what would be obtained via traditional methods (Fig. 3).**

@@

**Non-linear responses to climate variabels were common (Fig. 4).**
- Growth almost always increased with precipitation up to the long-term mean, but often declined under high precipitation.
- The most common response to temperture was unimodal, often peaking within 1SD of long-term mean. (**depending on how strongly this comes out, may warrent separate paper**)
- (results for cloudiness group)

@@

**Both $\Delta r$ and $\Delta AGB$ varied with $DBH$.** 
For $\Delta r$, the general tendency was a decline with DBH, sometimes following an initial increase (Fig. 2a). 
$\Delta AGB$ generally increased with $DBH$, although unimodal relationships were observed in some cases (Fig. 2b).

![**Figure 4 | Growth sensitivity to DBH:** (a) $\Delta r$, (b) $\Delta AGB$. THIS FIGURE IS JUST A SKETCH TO SHOW VALENTINE WHAT I HAVE IN MIND.](tables_figures/size_figure_sketch.png)

@@

@@

On DBH:
First, $\Delta r$ typically declines with tree size in cored individuals, consistent with many previous observations from tree-ring records (DENDRO_REFS).

DBH is not always collected when cores are taken, and is not routinely preserved alongside tree-ring data. For example, the International Tree-Ring Data Bank (ITRDB) contains no structure for storing DBH records.

>>>>

**Characterizing the influence of slowly-changing environtmental drivers is very challenging and uncertain.**
We want to know how changes in CO~2~ and atmospheric deposition are influencing tree growth.
A signficant challenge arises when tree-rings are used to characterize the influence of slowly changing environmental variables--climate, CO2, atmospheric deposition--on tree growth rates (REFS).
In these cases, a variety of methods are used to account for simultaneous changes in tree size, including... (DENDRO_REFS).
One commonly applied approach is to assume that $BAI$ is fundamentally independent of size (DENDRO_REFS).
None of these methods is fully satisfactory, leaving great uncertainty as to what the tree-ring record can tell us about the impact of rising CO~2~ (Walker et al. in revision). 
Rather, it is necessary to *simultaneously* account for the influences of changing tree size and environmental drivers.

**Here, develop a new method that allows simultaneous consideration of the effects of tree size, objectively determined principle climate drivers, and other environemntal drivers on annual tree growth.** 
This allows us to ask:
(1) What are the most important climate drivers of annual growth, and over which time windows?
(2) What is the shape of the relationship between annual growth and climate drivers?
(3) How do $\Delta r$, $BAI$, and aboveground biomass increments ($\Delta AGB$) vary with DBH?
(4) Are there interactions between DBH and climate drivers common?
(5) How is growth simultaneously influenced by DBH, climate, and other environemental drivers (CO~2~, atmospheric deposition)?

@@

First, we identified the most important climate driver for each of three categories of variables: a temperature group (mean, min, and max temperature; PET), a precipitation group (precipitation, number of days with precipitation), and a group linked to temperature variability and cloud cover (daily temperature range, percent cloud cover, and again PET). 
These groups were defined based on the dependencies of these variables on one another in the CRU database (Harris et al. 2014, [2020](https://rdcu.be/b3U3e)), as well as climatological and biological considerations.
Our first step was detrending to remove the influence of all non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), which is essential for identifying climatic drivers (DENDRO_REFS).
Specifically, we used a GAM model to fit a spline to individual tree growth records ($log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$), thereby producing residuals.
We note that an an attempt at the analysis without detrending failed to produce results consistent with biological expectations and results obtained via traditional dendrochronological methods, and generally failed to identify significant climate drivers.

We then used the *climwin* package in R (REF) [(van de Pol et. al. 2016)](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12590) to identify the most significant climate driver and time window for each of the three climate variable groups. 
*Climwin* searches through multiple climate variables and over a wide range of time frames to identify which are most strongly correlated to the biological variable of interest--in this case, residuals of $log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$.
Within *climwin*, we specified a mixed effects model using species and tree identity as random effect: `residual ~ [climate] + (1 | sp) + (1 | treeID)`. 
Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of time periods, at monthly resolution, starting in the previous May and ending in August of the current year.
We note that analyzing all species together yields the most significant climate drivers across the full set of cores from each site (our goal here), whereas identification of the top climate drivers for individual species would be optimized by analyzing each species separately.
*Climwin* can check for linear and quadratic relationships, and does k-fold cross-validation in its computation of AIC to guard against over-fitting (CLIMWIN_REF). 
We specified quadratic relationships only because (i) quadratic relationships are more consistent with known biological mechanisms (see Introduction), (ii) preliminary tests revealed that quadratic fits usually had lower AIC, and when not there tended to be little difference in AIC and the curve would approximate a straight line, (iii) modeling only quadratic improves coding efficiency and speeds up the analysis process.
For each group of candidate climate variables, we move forward with the best variable over the time window identified by climwin as a candidate climate variable for the multivariate models. 
When PET came out as the best variable in both temperature and cloud cover groups, there were only two candidate variables for the GAM. 
We checked for colinearity among the selected climate variables and removed any variable with a variance inflation factor > 10.

@@

A challenge to this system arose for the site undergoing most rapid climate change: Scotty Creek. 
There, [temperatures have increased by X $^\circ$ over X years]..., resulting in negative growth trends in basal area index (BAI) starting around 1950 and significant growth declines since 1970 in `r round(249/(249+69+127)*100, digits=0)`% of trees (Sniderhan & Baltzer 2016).
Problematically, correlating tree growth residuals from which climate-driven trends had been removed against the climate signal with a strong directional trend would not necessarily identify the most signficant climate drivers.
(**[ISSUE #25 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25): How do we solve this?**)

@@

Second, we combined DBH, climate, and atmospheric pollution data in a GLS model (Fig. 1).
[**DESCRIBE GLS**]

@@

**Our process picked out similar climate drivers to what would be obtained via traditional methods (Fig. 2).**

@@

**Non-linear responses to climate variabels were common (Fig. 3).**
- Growth almost always increased with precipitation up to the long-term mean, but often declined under high precipitation.
- The most common response to temperture was unimodal, often peaking within 1SD of long-term mean. (**depending on how strongly this comes out, may warrent separate paper**)
- (results for cloudiness group)

@@

** $\Delta r$, $BAI$, and $\Delta AGB$ all varied with $DBH$.** 
For $\Delta r$, the general tendency (with just a couple exceptions) was a decline with DBH, often following an initial increase (Fig. 4a). 
$BAI$ most commonly exibited a unimodal relationship (Fig. 4b).
$\Delta AGB$ also commonly exibited a unimodal relationship, although tending to peak at larger DBH (Fig. 4c).

![**Figure 4 | Growth sensitivity to DBH:** (a) $\Delta r$, (b) $BAI$, (c) $\Delta AGB$. THIS FIGURE IS JUST A ROUGH COMPILATION OF RESULTS SO FAR. Ideally, I'd like to plot all sites together, meaning we probably have to lose species IDs (plots with species IDs can be shown in appendix).](tables_figures/size_figure_sketch.png)

@@

We get different climate senstivity results with the different metrics of growth.

@@

**Our analysis of growth trends with DBH yields several novel insights for dendrochronology and forest ecology.**
First, the observation taht $\Delta r$ typically declines with tree size in cored individuals, often following initial increase during juvenile growth phase, is consistent with many previous observations from tree-ring records (DENDRO_REFS). 
This contrasts with what is seen at the stand level (Muller-Landau et al. 2006, Piponiot et al. in prep)-- presumably because [the sample of cored trees (survivors) doesn't match forest composition] (*cite paper that I reviewed several years back*).
Non-independence of $BAI$ from $DBH$ negates assumption that $BAI$ can be used as a DBH-independent metric of tree growth.
Our finding that $\Delta AGB$ is most commonly unimodal is surprising in that it contrasts with many previous findings and theoretical expectations (Stephenson et al, check Foster et al. 2006; check Piponiot et al. in prep). [EXPAND]
$\Delta AGB$ declines at high $DBH$ are presumably because trees are investing fixed C elsewhere--for example, reproduction.

DBH is not always collected when cores are taken, and is not routinely preserved alongside tree-ring data. For example, the International Tree-Ring Data Bank (ITRDB) contains no structure for storing DBH records. DEN is new alternative.
